\roottitle{Professional Experience} % Top level section

\headedsection{\href{https://www.tudublin.ie/research-innovation/research/discover-our-research/research-centres/centre-of-explainable-artificial-intelligence/}{Technological University Dublin}}
{\textsc{Dublin, Ireland}}
{

    \headedsubsection
    {External Scholar}
    {2025 - \phantom{20xx}}
    {
        \bodytext{
            at the Centre of eXplainable Artificial Intelligence. The Centre is the first of its kind in the Republic of Ireland and it aims to increase further and synergise cross-centres, college and external research collaboration.
        }
    }

}

\headedsection{Multidisciplinary Digital Publishing Institute (MDPI)}{}
{

    \headedsubsection
    {Topic Editor}
    {2024 - 2026}
    {
        \bodytext{
            for ``\href{https://www.mdpi.com/topics/TEYLN2388R}{Opportunities and Challenges in Explainable Artificial Intelligence}''.\\
            Submission pre-screening, review management and decision handling.
        }
    }

}


\headedsection{XAI4Science}{}
{

    \headedsubsection
    {Organizer}
    {2024 - 2025}
    {
        \bodytext{
            of the Workshop ``\href{https://xai4science.github.io/}{XAI4Science: From Understanding Model Behavior to Discovering\\ New Scientific Knowledge (2025)}'', co-located with ICLR 2025 at Singapore EXPO, Singapore.
        }
    }

}


\headedsection{World Conference on eXplainable Artificial Intelligence}{}
{

    \headedsubsection
    {Steering Committee Member}
    {2024 - \phantom{20xx}}
    {
        \bodytext{
            for the \href{https://xaiworldconference.com/2025/}{3rd XAI World Conference (2025)} in Istanbul, Turkey.\\
            Conference and Special Track co-organization.
        }
    }

    \headedsubsection
    {Programme Committee Chair}
    {2023 - 2024}
    {
        \bodytext{
            for the \href{https://xaiworldconference.com/2024/}{2nd XAI World Conference (2024)} in Valetta, Malta.\\
            Conference and Special Track co-organization.
        }
    }

}


\headedsection % Employer name which can include a hyperlink and location/URL on the right side of the page
{\href{https://www.hhi.fraunhofer.de/}{Fraunhofer Heinrich-Hertz-Institute}}%/\href{https://www.hhi.fraunhofer.de/}{\acr{HHI}}}
{\textsc{Berlin, Germany}}
{


    \headedsubsection % Job title entry for the current employer
    {Ethics Committee Member}
    {2023 - \phantom{20xx}}%{Jan '19 -- Dec '20}
    {
        \bodytext{
            Founding member of the first ethics committee at Fraunhofer HHI.
            

             \ifdefined\shortcv
               {}
             \else
               {

                \emph{Responsibilities:} Fulfilling the task and demand of assessing ethical aspects of research on and with humans, as well as general ethical issues.
                
                \vstep
               }
             \fi
        }
    }



    \headedsubsection
    {Head of Explainable Artificial Intelligence}
    {2021 - \phantom{20xx}}%{Jan '21 -- today}
    {
        \bodytext{
            Research Group Leadership and direction of XAI research \& applications.

            \ifdefined\shortcv
                {}
            \else
                {


                {\small(current number of staff: 2 PostDocs, 19 PhD researchers, 30+ student research assistants \& 3 Technical Staff).}
                \vstep

                \emph{Research:}
                Work towards the \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:u_35RYKgDlwC}{next generation of local-global Explainable AI} approaches
                and XAI-based model improvement by, e.g.,
                \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:Wp0gIr-vW9MC}{increasing efficiency}
                (\href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:TFP_iSt0sucC}{see also})
                and
                \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:iH-uZ7U-co4C}{debugging model training, reasoning and datasets}
                (\href{https://scholar.google.de/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:g5m5HwL7SMYC}{see also}).
                Provision of powerful modified backprop XAI for Pytorch models,
                and tools for reproducible XAI evaluations to the community, e.g., via
                \href{https://github.com/chr5tphr/zennit}{Zennit}, \href{https://github.com/rachtibat/LXT}{LXT}
                and
                \href{https://github.com/understandable-machine-intelligence-lab/Quantus}{Quantus}.

                \vstep

                \emph{Further responsibilities:}
                Project management and (funding) acquisition.
                Recruitment and guidance of research personnel.
                Founding member of the Ethics Committee at Fraunhofer HHI.
                }
            \fi

        }
    }

    \headedsubsection % Job title entry for the current employer
    {Tenured Researcher}
    {2019 - 2020}%{Jan '19 -- Dec '20}
    {
        \bodytext{
            PostDoc research position in the Machine Learning Group at Fraunhofer HHI.

             \ifdefined\shortcv
               {}
             \else
               {
                \emph{Research:}
                Development of \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:5nxA0vEk-isC}{Spectral Relevance Analysis},
                automating the detection of ``Clever Hans'' moments in machine learning.
                \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:3fE2CSJIrl8C}{Measurably increasing}
                the
                \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:MXK_kJrjxJIC}{explanation quality} of
                local XAI.
                Provision of modified backprop XAI in Keras/Tensorflow via
                \href{https://github.com/albermax/innvestigate}{iNNvestigate}.
    
                \vstep
    
                \emph{Further responsibilities:}
                Project (funding) acquisition.
                Recruitment and guidance of PhD students and student research assistants.
                
                \vstep
               }
             \fi




        }
    }

    \headedsubsection % Job title entry for the current employer
    {Research Associate}
    {2014 - 2018}%{Oct '14 -- Dec '18}
    {
        \bodytext{
            Founding member of the Machine Learning Group at Fraunhofer HHI.

            \ifdefined\shortcv
                {}
            \else
                {
                \vstep

                \emph{Research:}
                Furthering XAI research with the
                \href{https://scholar.google.com/citations?view_op=view_citation&hl=de&user=wpLQuroAAAAJ&citation_for_view=wpLQuroAAAAJ:u-x6o8ySG0sC}{development} and
                \href{https://scholar.google.com/citations?view_op=view_citation&hl=de&user=wpLQuroAAAAJ&citation_for_view=wpLQuroAAAAJ:u5HHmVD_uO8C}{evaluation} of corresponding methods,
                as well as applications in
                \href{https://scholar.google.com/citations?view_op=view_citation&hl=de&user=wpLQuroAAAAJ&citation_for_view=wpLQuroAAAAJ:9yKSN-GCB0IC}{various expert domains},
                resulting in
                \href{https://scholar.google.com/citations?user=wpLQuroAAAAJ}{several highly cited publications},
                open source
                \href{https://scholar.google.com/citations?view_op=view_citation&hl=de&user=wpLQuroAAAAJ&citation_for_view=wpLQuroAAAAJ:2osOgNQ5qMEC}{software}
                tools and
                \href{https://github.com/sebastian-lapuschkin/lrp_toolbox}{repositories},
                and the \href{https://scholar.google.com/citations?view_op=view_citation&hl=de&user=wpLQuroAAAAJ&citation_for_view=wpLQuroAAAAJ:d1gkVwhDpl0C}{first recorded encounter} of the
                \href{https://scholar.google.com/citations?view_op=view_citation&citation_for_view=wpLQuroAAAAJ:5nxA0vEk-isC}{``Clever Hans'' effect}
                in machine learning via XAI.

                \vstep

                \emph{Other contributions:}
                Contributions to the h.266 (VVC) video codec via learnable intra-frame prediction filters.
                Planning and conceptualization of an HPC cluster with modern GPU hardware implemented at Fraunhofer HHI.
                Development and showcasing
                \href{https://lrpserver.hhi.fraunhofer.de/}{multiple XAI demos}
                at international events.

                \vstep
                }
            \fi

          %Additional supervision by
          %  \href{https://scholar.google.com/citations?user=7aQwO08AAAAJ}{Prof. Dr. Wojciech Samek}.

        }
    }
}

%------------------------------------------------

\headedsection % Employer name which can include a hyperlink and location/URL on the right side of the page
{\href{http://www.tu-berlin.de/ }{Berlin Institute of Technology}}%/\href{http://www.tu-berlin.de/ }{\acr{TU Berlin}}}
{\textsc{Berlin, Germany}}
{
    \headedsubsection % Job title entry for the current employer
    {Research Associate}
    {2013 - 2014}%{Sep '13 -- Sep '14}
    {
        \bodytext{

            \ifdefined\shortcv
                {}
            \else
                {
                    \emph{Research:}
                    Formalization and development of the ``\href{https://scholar.google.com/citations?view_op=view_citation&hl=de&user=wpLQuroAAAAJ&citation_for_view=wpLQuroAAAAJ:YsMSGLbcyi4C}{Layer-wise Relevance Propagation}'' (LRP) method
                    of Explainable AI
                    for explaining individual predictions of nonlinear machine learning models.

                    \vstep
                }
            \fi

            Supervision by
            \href{https://scholar.google.com/citations?user=jplQac8AAAAJ}{Prof. Dr. Klaus-Robert M\"uller} and
            \href{https://scholar.google.com/citations?user=5B8CTlEAAAAJ}{Prof. Dr. Alexander Binder}.

        }
    }

    \headedsubsection % Job title entry for the current employer
    {Student Research- \& Teaching Assistant}
    {2011 - 2013}%{Oct '11 -- Aug '13}
    {
        \bodytext{
            \ifdefined\shortcv
                {}
            \else
                {
                    \emph{Research:}
                    Structure and cell type detection in large histopathology images using Bag of Words features and SVM classifiers.
                    Development of XAI for the pipeline.
                }
            \fi

            Research assistant to
            \href{https://scholar.google.com/citations?user=5B8CTlEAAAAJ}{Prof. Dr. Alexander Binder}
            at the machine learning group at TU Berlin.


            \ifdefined\shortcv
                {}
            \else
                {
                    \vstep

                    \emph{Teaching:}
                    Preparation and lecturing (of exercise sessions) in the courses
                    ``Machine Learning~1'' and ``Machine Learning~2 -- Theory and Application''  and associated academic courses.
                    Visualization and animation of data and learning algorithms discussed throughout the course work.

                    \vstep
                }
            \fi

            Teaching assistant to
            \href{https://scholar.google.com/citations?user=jplQac8AAAAJ}{Prof. Dr. Klaus-Robert M\"uller},
            \href{https://scholar.google.com/citations?user=VYi_04kAAAAJ}{Prof. Dr. Dr. Franz Kir\'aly},\\
            \href{https://scholar.google.com/citations?user=akYb0wsAAAAJ}{Dr. Irene Dowding (n\'ee Winkler)} and
            \href{https://www.linkedin.com/in/bartzdaniel}{Dr. Daniel Bartz}.
        }
    }

    \headedsubsection % Job title entry for the current employer
    {Student Teaching Assistant}
    {2009 - 2011}%{Oct '09 -- Sep '11}
    {
        \bodytext{
            \ifdefined\shortcv
                {}
            \else
                {
                    Course instruction for algorithmic and practical foundations of computer science (B.Sc.):
                    Basic and advanced Java development, software engineering and OOP concepts, algorithms on image and graph data, among others.

                    \vstep
                }
            \fi

            Teaching assistant to
            \href{https://scholar.google.com/citations?user=KNwXLLsAAAAJ}{Prof. Dr. Marc Alexa},
            \href{https://scholar.google.com/citations?user=mvOSfFAAAAAJ}{Prof. Dr. Odej Kao} and
            \href{https://scholar.google.com/citations?user=83OxU_4AAAAJ}{Prof. Dr. Oliver Brock}.

        }
    }
}
